Liveness-Based Garbage Collection for Lazy Languages

### 2016-02-13 paper/paper.tex
proof writing

\definecolor{Myblue}{rgb}{.2,0,1}
\definecolor{Myred}{rgb}{1,.2,.2}
\newcommand{\mycomment}[1]{{\color{Myblue}{#1}}}
\newcommand{\added}[1]{{\color{Myblue}{#1}}}
%\newcommand{\added}[1]{#1}
\newcommand{\todelete}[1]{{\color{Myred}{#1}}}

@@

We next  describe the function $\mathcal{L}$  that propagates liveness
across  expressions.   Consider  the $\mathcal{L}$-rules  for  {\LET},
{\SIF}, and {\SRETURN}.  Since the value of $(\SRETURN~x)$ is the
value  of  $x$,  a  demand $\sigma$  on  $(\SRETURN~x)$  gives  a
liveness   of  $\{x.\sigma\}$.    The  liveness   of  the   expression
$(\SIF~x~e_1~e_2)$  is  a union  of  the  liveness of  $e_1$  and
$e_2$. In  addition, since  the condition $x$  is also  evaluated, the
liveness $\{x.\epsilon\}$ is created and  added to the union.  
To  understand  the  liveness rule  for  $\LET~x  \leftarrow~s~\IN~e$,
observe that the value  of $\LET$ is the value of  its body $e$.  Thus
the  liveness environment  $\Lv$ of  $e$ is  calculated for  the given
demand $\sigma$. Since the stack variable $x$ is copied to each of the
closure variable  $x_{\pi}$, the liveness of  $x$ is the union  of the
liveness of the  closure variables.  This liveness,  say $\sigma'$, is
also  the  demand  on  $s$, thus the  liveness  environment
$\mathit{ref}(s,   \sigma',   \Lfonly)$   is  added   to   $\Lv   \cup
\{x.\sigma\}$. Finally, the stack variables corresponding to the free
variables of $s$ are updated and added 
  to  give the  overall liveness environment  for $(\LET~x
\leftarrow~s~\IN~e)$.
{\color{Myblue}
As noted earlier,  $x.\alpha \in \Lv_y$ specifies the  liveness of the
reference $\heap\llbracket\rho(x), \alpha\rrbracket$  only if $\alpha$
corresponds to a closure-free path  in \heap\ starting from $\rho(x)$.
If this path is intercepted by a closure, say $(\CAR~{x_{\pi}})$, then
the liveness of heap contents starting from $\mathit{x_{\pi}}$
is given by $\Lv_\mathit{x_{\pi}}$.

 As we  shall see in section??,  the liveness of the  closure variable
 $x_{\pi}$ is recorded along with the closure so that the GC can refer
 to it during garbage collection.
}
%} %comment

@@

We shall now present a proof of soundness of the analysis presented in
Section~\ref{sec:liveness-analysis}.   Here are  the ideas  behind the
proof.
\begin{enumerate}
\item      We     augment      the      standard     semantics      in
  Figure~\ref{fig:lang-semantics} to  model a GC before  the execution
  of  each  \LET\  expression.   {\color{Myblue}Note  that  memory  is
    allocated only during execution  of \LET\ expressions}. During GC,
  we  track each  reference  in the  root  set and  heap  that is  not
  declared  live by  our analysis.   Any attempt  to dereference  such
  references results in the transition system entering a special state
  denoted   \bang.   We   call  the   semantics  after   augmentation,
  \emph{minefield semantics}.

%% During GC, we bind  a special value
%%   $\bot$ to each  reference in the root  set and heap  that is not
%%   live by  our analysis.  This  models the de-allocation  of locations
%%   without any  live references.  Any  attempt to dereference  a $\bot$
%%   during execution results
%% %\added{that  has   been   declared  not   live  by   our
%% %    analysis} %%with  a null-liveness during a  transition will result
%%   in the transition system entering  a special state denoted \bang.  We
%%   call the semantics after augmentation, \emph{minefield semantics}.
\item \label{inline1} Assuming that a program enters the \bang\ state,
  we  show  how to  construct,  through  inline expansion,  a  program
  without function  calls which has  the same minefield  behavior. The
  final step shows that no program  without function calls can enter the
  \bang\ state. As a consequence  no program (with or without function
  calls) can enter the \bang\ state.
\end{enumerate}

@@

\hline 
\makecell[t]{ $\rho(x) \mbox{ is } \bot$} & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$ &
{\sc car-bang} 
\\
\hline
\makecell[t]{$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$} & \makecell[t]{$\rho,\, S,\,  \heap,\,
(\CAR~x), \cred{\sigma} \rightsquigarrow$ \\ $ \rho', \,(\rho, addr(\langle
s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$ }     &
{\sc car-1-clo} 
\\
\hline
\makecell[t]{ $\heap(\rho(x))
\mbox{ is } \langle s, \rho'\rangle$} & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, \rho(x),
  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, 
\cred{\renewcommand{\arraystretch}{1}\left\{
\begin{array}{lr}
  {\emptyset}&\mbox{if } \sigma = \emptyset \\       
  {\{\epsilon\} \cup \acar\sigma }&\mbox{otherwise}
\end{array}\right.
}$         
&
{\sc car-clo}\\
\hline
\end{tabular}
\end{center}
\caption{Minefield semantics for {\sc car-bang} and
  {\sc car-clo} \warning{Add a note mentioning that we handle the issue of requiring a specific demand by introducing a new symbol}\label{fig:minefield-semantics-for-some}}
\end{figure*}

@@

\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness-based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to non-live references being replaced by $\bot$.  This simulates the
  act of  not retaining  (effectively garbage collecting)  the cells
  pointed to by these references  during an actual GC.
  This is done in the following way:
 \begin{enumerate}

 \item  Consider  the following  environments,  1)  environment in  the
  current state, 2)  environments in each of the stacked  items and 3)
  the environment  in each of  the closures in  the heap. For  each of
  these  environments  we  can   calculate  a  corresponding  liveness
  environment using the $e\ and\ \sigma$.

  \item {\color{Myblue}\emph{GC} for  each $x \in domain(\rho)$, we  set $\rho(x)$ to
    $\bot$  iff $x.\epsilon  \notin \Lv$.   Similarly, for  each stack
    entry   $(\rho,\ell,e,\sigma)$  in   $S$   with   $\Lv'$  as   the
    corresponding liveness  environment in $\mathsf{SL}$,  and finally
    for  each  $x  \in  domain(\rho)$ for  some  closure  $\langle  s,
    \lfloor\rho\rfloor_{FV(s)}, \sigma\rangle$ in the heap, for all $x
    \in  FV(s)$ we  set $x$  to  $\bot$ iff  $x.\epsilon \notin  \Lv'$.}

  \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
    \sigma$  and the  demand  transformers $\Lfonly$,  we construct  a
    liveness environment  \Lv\ for  the expression $e$.   We similarly
    define the liveness  environments for the expressions  $e$ in each
    of  the continuation  frames  $(\rho, \ell,  e,  \sigma)$ in  $S$,
    giving  a stack  of  liveness environments  that  we shall  denote
    $\mathsf{SL}$.
  \item  \emph{Garbage collection  of objects  pointed by  root sets:}
    Consider  each   of  the  following  environments:   (i)  The  the
    environment $\rho$ in  the current state (ii) and for  each $x \in
    domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
    \Lv$.  Similarly,  for each stack entry  $(\rho,\ell,e,\sigma)$ in
    $S$  with  $\Lv'$ as  the  corresponding  liveness environment  in
    $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
    to $\bot$ iff $x.\epsilon \notin \Lv'$.

Consider each of  the following environments: (i)  The the environment
$\rho$ in the current state and  for each $x \in domain(\rho)$, we set
$\rho(x)$ to $\bot$ iff $x.\epsilon  \notin \Lv$.  Similarly, for each
stack  entry   $(\rho,\ell,e,\sigma)$  in  $S$  with   $\Lv'$  as  the
corresponding liveness  environment in $\mathsf{SL}$, and  for each $x
\in domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
\Lv'$.

  \item \emph{Garbage  collection of  objects in  the heap:}

    For each
    location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
    environment $\rho$ in either  the  current state or in a
    continuation frame, for each $x \in domain(\rho)$ and each forward
    access
    path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
    \rrbracket = \ell$. Here 
%% $x.\alpha$  is  live  and
%%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
%%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
\Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
every transition  is preceded  by a  GC
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in
Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
earlier  call to  $GC$  results  in $\rho(x)$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-bang}). Otherwise
the transition is similar to the earlier {\sc car-clo}
rule.

@@

\makecell[t]{$\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap,
  (+~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $\rho', S, \heap[\ell
    := \heap(\rho'(x)) + \heap(\rho'(y))], e, \cred{\sigma'}$}     
 &
{\sc prim-add} \\
\hline
\makecell[t]{$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$} &\makecell[t]{$\rho, S,
  \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $ \rho', (\rho, \rho(x),
  (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
{\sc prim-1-clo} \\
\hline
\makecell[t]{$\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $} & \makecell[t]{$\rho,
  S, \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $ \rho', (\rho,\rho(y),
  (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
{\sc prim-2-clo} \\
\hline
\makecell[t]{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & \makecell[t]{$\rho, S, \heap,
  (f~\myvec{x}), \cred{\sigma}$  $\rightsquigarrow$\\$ [\myvec{y} \mapsto
    \rho(\myvec{x})], S, \heap, e_{\mathit{f}}, \cred{\sigma}$}      &
{\sc funcall} \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S, \heap_1, (\LET~x\leftarrow
  s~\IN~e), \sigma) = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow$ \\ $ \rho\oplus[x
    \mapsto \ell], S, \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}, \sigma_x\rangle], e, \cred{\sigma}$ \\
    where $\sigma_x\ =\  \lfloor\mathcal{L}(e,\sigma,\Lfonly)\rfloor_{\{x\}}$} &
{\sc let} \\ 
\hline
\makecell[t]{ $\rho(x) \mbox{ is } \bot$} & $\rho, S,
  \heap, (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma} \rightsquigarrow \bang$   &
{\sc if-bang} 
\\
\hline
\makecell[t]{$\heap(\rho(x)) \ne 0$}  & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
  \cred{\sigma}$  $\rightsquigarrow \rho, S, \heap,  e_1, \cred{\sigma}$} & {\sc if-true} \\
\hline
\makecell[t]{$\heap(\rho(x)) = 0$} & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
  \cred{\sigma}$   $\rightsquigarrow
\rho, S, \heap,  e_2, \cred{\sigma}$} & {\sc if-false} \\
\hline
\makecell[t]{$\heap(\rho(x)) = \langle s, \rho' \rangle $} & \makecell[t]{$\rho, S, \heap,
  (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma}$ $\rightsquigarrow$ \\ $
\rho', (\rho, \rho(x), (\SIF~x~e_1~e_2),  \cred{\sigma})\!:\!S, \heap, s,
\cred{\clazy\sigma}$}
&
{\sc if-clo} \\
\hline
\makecell[t]{
  $\rho(x) \mbox{ is } \bot$} & $\rho, S,
  \heap, (\SRETURN~x), \cred{\sigma} \rightsquigarrow \bang$   &
{\sc return-bang} 
\\
\hline
\makecell[t]{$\heap(\rho(x))~\mbox{is in WHNF with
    value}~v$} & \makecell[t]{$\rho,
  (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap, (\SRETURN~x), \cred{\sigma}$  $\rightsquigarrow$\\$ \rho', S, \heap[\ell := v], e, 
  \cred{\sigma'}$} &
{\sc return-whnf}\\
\hline
\makecell[t]{$\heap(\rho(x)) = \langle s, \rho' \rangle $} & \makecell[t]{$\rho, S, \heap,
  (\SRETURN~x), \cred{\sigma}$  $
  \rightsquigarrow$ \\
$\rho',~ (\rho, \rho(x), (\SRETURN~x), \cred{\sigma})\!:\!S, \heap,  s,
  \cred{\sigma}$} &
{\sc return-clo} \\
\hline
\end{tabular}

>>>>

\definecolor{Myblue}{rgb}{.2,0,1}
\definecolor{Myred}{rgb}{1,.2,.2}
\newcommand{\mycomment}[1]{{\color{Myblue}{#1}}}
\newcommand{\added}[1]{{\color{Myblue}{#1}}}
%\newcommand{\added}[1]{#1}
\newcommand{\todelete}[1]{}%{\color{Myred}{#1}}}

@@

We next  describe the function $\mathcal{L}$  that propagates liveness
across  expressions.   Consider  the $\mathcal{L}$-rules  for  {\LET},
{\SIF}, and {\SRETURN}.  Since the value of $(\SRETURN~x)$ is the
value  of  $x$,  a  demand $\sigma$  on  $(\SRETURN~x)$  gives  a
liveness   of  $\{x.\sigma\}$.    The  liveness   of  the   expression
$(\SIF~x~e_1~e_2)$  is  a union  of  the  liveness of  $e_1$  and
$e_2$. In  addition, since  the condition $x$  is also  evaluated, the
liveness $\{x.\epsilon\}$ is created and  added to the union.  
To  understand  the  liveness rule  for  $\LET~x  \leftarrow~s~\IN~e$,
observe that the value  of $\LET$ is the value of  its body $e$.  Thus
the  liveness environment  $\Lv$ of  $e$ is  calculated for  the given
demand $\sigma$. Since the stack variable $x$ is copied to each of the
closure variable  $x_{\pi}$, the liveness of  $x$ is the union  of the
liveness of the  closure variables.  This liveness,  say $\sigma'$, is
also  the  demand  on  $s$, thus the  liveness  environment
$\mathit{ref}(s,   \sigma',   \Lfonly)$   is  added   to   $\Lv   \cup
\{x.\sigma\}$. Finally, the stack variables corresponding to the free
variables of $s$ are updated and added 
  to  give the  overall liveness environment  for $(\LET~x
\leftarrow~s~\IN~e)$.

{\color{Myblue}  As noted  earlier, $x.\alpha  \in \Lv$  specifies the
  liveness of the reference $\heap\llbracket\rho(x), \alpha\rrbracket$
  only   if   $\alpha$  corresponds   to   a   closure-free  path   in
  \heap\ starting  from $\rho(x)$.  If  this path is intercepted  by a
  closure,  say  $(\CAR~{y_{\pi}})$, then  the  liveness  of the  path
  starting from $\mathit{y_{\pi}}$ is given by $\Lv_\mathit{y_{\pi}}$.
  As we shall see in  Section~\ref{sec:GC-scheme}, the liveness of the
  closure variable  $y_{\pi}$ is recorded  along with the  closure for
  $s$ so that the GC can refer to it during garbage collection.  }
%} %comment

@@

We shall now present a proof of soundness of the analysis presented in
Section~\ref{sec:liveness-analysis}. {\color{Myblue} It is easy to see that the analysis
correctly identifies the liveness of stack variables. A stack variable
is live
between its introduction through a \LET\  and its last use to
create a closure variable. This is correctly captured by the
\LET\ rule in Figure~\ref{fig:live-judge}.}

Proving soundness  for closure  variables is more complex. Here  are the
ideas behind the proof. 
\begin{enumerate}
\item      We     augment      the      standard     semantics      in
  Figure~\ref{fig:lang-semantics} to  model a GC before  the execution
  of  each \LET\  expression.  {\color{Myblue}Note  that, unlike  eager
    languages,   memory  is   allocated  only   during  execution   of
    \LET\ expressions}. During GC, we track each reference in the root
  set and heap that is declared  dead by our analysis.  Any attempt to
  dereference  such  references  results   in  the  transition  system
  entering a special state denoted \bang.  We call the semantics after
  augmentation, \emph{minefield semantics}.

%% During GC, we bind  a special value
%%   $\bot$ to each  reference in the root  set and heap  that is not
%%   live by  our analysis.  This  models the de-allocation  of locations
%%   without any  live references.  Any  attempt to dereference  a $\bot$
%%   during execution results
%% %\added{that  has   been   declared  not   live  by   our
%% %    analysis} %%with  a null-liveness during a  transition will result
%%   in the transition system entering  a special state denoted \bang.  We
%%   call the semantics after augmentation, \emph{minefield semantics}.
\item \label{inline1} Assuming that a program enters the \bang\ state,
  we  construct,  through  inline expansion,  a  program
  without function  calls which has  the same minefield  behavior. The
  final step shows that no program  without function calls can enter the
  \bang\ state. As a consequence  no program (with or without function
  calls) can enter the \bang\ state.
\end{enumerate}

@@

\hline 
\makecell[t]{ $\rho(x) \mbox{ is } \bot$} & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow \bang$ &
{\sc car-bang} 
\\
\hline
\makecell[t]{$\heap(\rho(x)) \mbox{ is } (\langle s, \rho'\rangle, d)$} & \makecell[t]{$\rho,\, S,\,  \heap,\,
(\CAR~x), \cred{\sigma} \rightsquigarrow$ \\ $ \rho', \,(\rho, addr(\langle
s, \rho'\rangle), (\CAR~x),\cred{\sigma} )\!:\!S,\, \heap,\, s, \, \cred{\sigma}$ }     &
{\sc car-1-clo} 
\\
\hline
\makecell[t]{ $\heap(\rho(x))
\mbox{ is } \langle s, \rho'\rangle$} & $\rho, S,
  \heap, (\CAR~x), \cred{\sigma} \rightsquigarrow   \rho', (\rho, \rho(x),
  (\CAR~x), \cred{\sigma})\!:\!S, \heap, s, 
\cred{\renewcommand{\arraystretch}{1}\left\{
\begin{array}{lr}
  {\emptyset}&\mbox{if } \sigma = \emptyset \\       
  {\{\epsilon\} \cup \acar\sigma }&\mbox{otherwise}
\end{array}\right.
}$         
&
{\sc car-clo}\\
\hline
\makecell[t]{$\cred{GC(\rho_1, S_1, \heap_1, (\LET~x\leftarrow
  s~\IN~e), \sigma) = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow$ \\ $ \rho\oplus[x
    \mapsto \ell], S, \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}, \sigma_x\rangle], e, \cred{\sigma}$ \\
    where $\sigma_x\ =\  \lfloor\mathcal{L}(e,\sigma,\Lfonly)\rfloor_{\{x\}}$} &
{\sc let} \\ 
\hline
\end{tabular}
\end{center}
\caption{Minefield  semantics for  {\sc  car-bang}  and {\sc  car-clo}
  \warning{Add a note mentioning that we handle the issue of requiring
    a      specific     demand      by      introducing     a      new
    symbol}\label{fig:minefield-semantics-for-some}}
\end{figure*}

@@

\item  $GC(\rho,  S,  \heap,  e, \sigma)$ models  a  liveness-based  garbage
  collection that returns $(\rho', S', \heap')$, where the changes in
  $\rho, S$ and  $\heap$ are due
  to non-live references being replaced by $\bot$.  This simulates the
  act of  garbage collecting  the cells
  pointed to by these references  during an actual garbage collection.
  To do so, $GC$ needs to consider the following environments:  (1) the  environment in  the
  current state, (2) the  environment in each of the stacked  continuations and (3)
  the environment  in each of  the closures in  the heap.

 \begin{enumerate} 
{\color{Myblue}

 \item \label{env}  For each of  these environments, $GC$  calculates a
   liveness  environment $\Lv$  using  the  corresponding  $e$  (or  $s$)  and
   $\sigma$.
 \item 
   For each
    location $\ell$,  $GC$  sets $\heap(\ell)$  to $\bot$ iff for each
    environment $\rho$ above, for each $x \in domain(\rho)$, and each forward
    access
    path $\alpha$, it is \emph{not} the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
    \rrbracket = \ell$. 
%% $x.\alpha$  is  live  and
%%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
%%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
}

  %% \item {\color{Myblue}\emph{GC} for  each $x \in domain(\rho)$, we  set $\rho(x)$ to
  %%   $\bot$  iff $x.\epsilon  \notin \Lv$.   Similarly, for  each stack
  %%   entry   $(\rho,\ell,e,\sigma)$  in   $S$   with   $\Lv'$  as   the
  %%   corresponding liveness  environment in $\mathsf{SL}$,  and finally
  %%   for  each  $x  \in  domain(\rho)$ for  some  closure  $\langle  s,
  %%   \lfloor\rho\rfloor_{FV(s)}, \sigma\rangle$ in the heap, for all $x
  %%   \in  FV(s)$ we  set $x$  to  $\bot$ iff  $x.\epsilon \notin  \Lv'$.}

  %% \item  \label{gc-1}Using the  current state  $\rho, \stk,  \heap, e,
  %%   \sigma$  and the  demand  transformers $\Lfonly$,  we construct  a
  %%   liveness environment  \Lv\ for  the expression $e$.   We similarly
  %%   define the liveness  environments for the expressions  $e$ in each
  %%   of  the continuation  frames  $(\rho, \ell,  e,  \sigma)$ in  $S$,
  %%   giving  a stack  of  liveness environments  that  we shall  denote
  %%   $\mathsf{SL}$.
  %% \item  \emph{Garbage collection  of objects  pointed by  root sets:}
  %%   Consider  each   of  the  following  environments:   (i)  The  the
  %%   environment $\rho$ in  the current state (ii) and for  each $x \in
  %%   domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
  %%   \Lv$.  Similarly,  for each stack entry  $(\rho,\ell,e,\sigma)$ in
  %%   $S$  with  $\Lv'$ as  the  corresponding  liveness environment  in
  %%   $\mathsf{SL}$, and for each $x \in domain(\rho)$, we set $\rho(x)$
  %%   to $\bot$ iff $x.\epsilon \notin \Lv'$.

%% Consider each of  the following environments: (i)  The the environment
%% $\rho$ in the current state and  for each $x \in domain(\rho)$, we set
%% $\rho(x)$ to $\bot$ iff $x.\epsilon  \notin \Lv$.  Similarly, for each
%% stack  entry   $(\rho,\ell,e,\sigma)$  in  $S$  with   $\Lv'$  as  the
%% corresponding liveness  environment in $\mathsf{SL}$, and  for each $x
%% \in domain(\rho)$, we  set $\rho(x)$ to $\bot$  iff $x.\epsilon \notin
%% \Lv'$.

%%   \item \emph{Garbage  collection of  objects in  the heap:}

%%     For each
%%     location $\ell$,  we set $\heap(\ell)$  to $\bot$ iff for each
%%     environment $\rho$ in either  the  current state or in a
%%     continuation frame, for each $x \in domain(\rho)$ and each forward
%%     access
%%     path $\alpha$, it is not the case that $x.\alpha \in \Lv$ and $\heap\llbracket\rho(x), \alpha
%%     \rrbracket = \ell$. Here 
%% %% $x.\alpha$  is  live  and
%% %%     corresponds  to   $\ell$ in \heap.  More  formally,  there   are  no  $x\in
%% %%     domain(\rho)$ and  $\alpha$, such that  $\heap\llbracket\rho(x), \alpha
%% %%     \rrbracket = \ell$  and $x.\alpha \in \Lv$, where 
%% \Lv\ is the liveness environment  corresponding to $\rho$ mentioned in
%% step~\ref{gc-1}.
  \end{enumerate}
\end{enumerate}
Starting from the initial state,
%%  $([\;]_\rho,([\;]_\rho,
%% \ans, (\print~\ans)):[\;]_{S}  , [\;]_{\heap}, (\mainpgm)$,
every transition  is preceded  by a  GC
using $GC$.  The details of the transition for the {\sc
  car-clo}  rule   is  shown   in
Figure~\ref{fig:minefield-semantics-for-some}.  If  this  or  an
earlier  call to  $GC$  results  in $\rho(x)$  being
bound  to  $\bot$,  then  the  $\rightsquigarrow$  step
enters the  \bang\ state ({\sc  car-bang}). Otherwise
the transition is similar to the earlier {\sc car-clo}
rule.

@@

\makecell[t]{$\rho, (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap,
  (+~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $\rho', S, \heap[\ell
    := \heap(\rho'(x)) + \heap(\rho'(y))], e, \cred{\sigma'}$}     
 &
{\sc prim-add} \\
\hline
\makecell[t]{$\heap(\rho(x)) \mbox{ is } \langle s, \rho'\rangle$} &\makecell[t]{$\rho, S,
  \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $ \rho', (\rho, \rho(x),
  (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
{\sc prim-1-clo} \\
\hline
\makecell[t]{$\heap(\rho(y)) \mbox{ is } \langle s, \rho'\rangle $} & \makecell[t]{$\rho,
  S, \heap, (+~x~y), \cred{\sigma}$  $\rightsquigarrow$ \\ $ \rho', (\rho,\rho(y),
  (+~x~y), \cred{\sigma})\!:\!S, \heap, s, \cred{\clazy\sigma}$}      &
{\sc prim-2-clo} \\
\hline
\makecell[t]{$\mathit{f}~\mbox{defined as}$
$~(\DEFINE~(f~\myvec{y})~e_{\mathit{f}})$}  & \makecell[t]{$\rho, S, \heap,
  (f~\myvec{x}), \cred{\sigma}$  $\rightsquigarrow$\\$ [\myvec{y} \mapsto
    \rho(\myvec{x})], S, \heap, e_{\mathit{f}}, \cred{\sigma}$}      &
{\sc funcall} \\
\hline
\makecell[t]{$\cred{GC(\rho_1, S_1, \heap_1, (\LET~x\leftarrow
  s~\IN~e), \sigma) = (\rho, S, \heap)}$,\\$\ell$ is a new location}& \makecell[t]{$\rho, S, \heap, (\LET~x\leftarrow
  s~\IN~e), \cred{\sigma}$  $ \rightsquigarrow$ \\ $ \rho\oplus[x
    \mapsto \ell], S, \heap[\ell := \langle s, \lfloor\rho\rfloor_{FV(s)}, \sigma_x\rangle], e, \cred{\sigma}$ \\
    where $\sigma_x\ =\  \lfloor\mathcal{L}(e,\sigma,\Lfonly)\rfloor_{\{x\}}$} &
{\sc let} \\ 
\hline
\makecell[t]{ $\rho(x) \mbox{ is } \bot$} & $\rho, S,
  \heap, (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma} \rightsquigarrow \bang$   &
{\sc if-bang} 
\\
\hline
\makecell[t]{$\heap(\rho(x)) \ne 0$}  & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
  \cred{\sigma}$  $\rightsquigarrow \rho, S, \heap,  e_1, \cred{\sigma}$} & {\sc if-true} \\
\hline
\makecell[t]{$\heap(\rho(x)) = 0$} & \makecell{$\rho, S, \heap, (\pi:\SIF~\psi:x~e_1~e_2),
  \cred{\sigma}$   $\rightsquigarrow
\rho, S, \heap,  e_2, \cred{\sigma}$} & {\sc if-false} \\
\hline
\makecell[t]{$\heap(\rho(x)) = \langle s, \rho' \rangle $} & \makecell[t]{$\rho, S, \heap,
  (\pi:\SIF~\psi:x~e_1~e_2), \cred{\sigma}$ $\rightsquigarrow$ \\ $
\rho', (\rho, \rho(x), (\SIF~x~e_1~e_2),  \cred{\sigma})\!:\!S, \heap, s,
\cred{\clazy\sigma}$}
&
{\sc if-clo} \\
\hline
\makecell[t]{
  $\rho(x) \mbox{ is } \bot$} & $\rho, S,
  \heap, (\SRETURN~x), \cred{\sigma} \rightsquigarrow \bang$   &
{\sc return-bang} 
\\
\hline
\makecell[t]{$\heap(\rho(x))~\mbox{is in WHNF with
    value}~v$} & \makecell[t]{$\rho,
  (\rho', \ell, e, \cred{\sigma'})\!:\!S, \heap, (\SRETURN~x), \cred{\sigma}$  $\rightsquigarrow$\\$ \rho', S, \heap[\ell := v], e, 
  \cred{\sigma'}$} &
{\sc return-whnf}\\
\hline
\makecell[t]{$\heap(\rho(x)) = \langle s, \rho' \rangle $} & \makecell[t]{$\rho, S, \heap,
  (\SRETURN~x), \cred{\sigma}$  $
  \rightsquigarrow$ \\
$\rho',~ (\rho, \rho(x), (\SRETURN~x), \cred{\sigma})\!:\!S, \heap,  s,
  \cred{\sigma}$} &
{\sc return-clo} \\
\hline
\end{tabular}
